% !TEX root = CSC104LectureNotes.tex

\chapter{About These Notes}

\section*{Welcome Future Programmer!}

Hello!  Thank you and congratulations for choosing to study programing and problem solving at Nassau Community College!  Computers are ubiquitous in our society, and so a knowledge of how they work is very important for the educated person in the 21st century.  This course will provide a unique opportunity for you to discover how computers work, by telling them what to do!

While this course will not make you an expert programmer, learning the basics of programming -- and, perhaps just as importantly, learning how programmers think -- will give you the insights you need in understand what drives our economy, our culture, our technology, and virtually every other facet of our lives.  Even a cursory understanding of programming and of programmers will make you a better physicist, or accountant, or elementary school teacher.

\section*{How to Use This Document}

This document is meant to introduce you to the topics that you will be covering each day in class.  It is recommended that you read each day's notes \textbf{\textit{before}} coming to class.  The discussions you have in class are not meant to introduce you to these topics, but to \textbf{\textit{reinforce}} the topics.  The best strategy for succeeding in this class is to avoid surprises!  Arm yourself with as much information as possible \textit{before} attending class, and then capitalize on your new knowledge after class by \textit{practicing} what you've learned.

\subsection*{Ethical Behavior}

As Stan Lee taught us through his character Spider-Man, with great power comes great responsibility.  As the world's activities become increasingly electronic, and increasingly interconnected, we programmers must do our part to ensure the rights -- including the right to privacy -- of those whose lives our work will affect.  The ethics of programming will be discussed at various points in these notes and during the course.

To that end, we NCC Computer Science and Information Technology faculty subscribe to the \href{https://www.acm.org/code-of-ethics}{Code of Ethics} published by the \href{https://www.acm.org/}{Association for Computing Machinery}, and we will refer to this Code from time to time.

\section*{How This Document Was Created}

The tools used to create this document include:

\bi
    \item The \LaTeX{} document typesetting system (\url{https://www.latex-project.org/})
    \bi
        \item The \hologo{XeLaTeX} (\url{http://xetex.sourceforge.net/}) typesetting extension
    \ei
    \item Mozilla Project's Fira Type Family (\url{https://github.com/mozilla/Fira})
    \item VS Code (\url{https://code.visualstudio.com/})
    \bi
        \item James Yu's LaTeX Workshop extension (\url{https://marketplace.visualstudio.com/items?itemName=James-Yu.latex-workshop})
    \ei
    \item Green Mountain Coffee Roasters Colombia Select (\url{http://www.gmcr.com/})
\ei